---
title: "STA330_ScRNA_PBMCs_10xGenomics"
author: "Duy Nguyen, Minh Nguyen, Hannah Kim"
date: "2025-02-05"
output: pdf_document
---

### 0) General Notes

scRNA-Seq enables high-resolution, genome-wide profiling of RNA molecules in individual
cells. A primary application is assessing transcriptional similarities and differences 
within cell populations, revealing significant heterogeneity in contexts such as embryonic
and immune cells. This heterogeneity analysis remains central to scRNA-Seq studies. In 
addition, scRNA-Seq helps identify rare cell populations - e.g., malignant tumor cells 
within a tumor mass or hyper-responsive immune cells—that would be undetectable in bulk 
analyses. 

Beyond cellular heterogeneity, scRNA-Seq provides insights into gene expression 
fundamentals, including monoallelic expression, splicing patterns, and transcriptional 
noise. It can also identify co-regulated gene modules and infer gene-regulatory networks 
that drive functional heterogeneity and cell-type specification.

However, the level of detail scRNA-Seq provides depends on the specific protocol used.

The general procedure for a scRNA-Seq experiment

1. Experimental procedure

- Isolate single cells from a tissue sample

- Single cell lysis in a way that preserve the cellular mRNA

- mRNA molecule captures using poly-T sequence primers that bind to mRNA poly-A
tails 

- Convert the poly-T-primed mRNA into cDNA using reverse transcription

- cDNA amplification via PCR or in vitro transcription 

- cDNA sequencing library preparation 

- Pool cDNA sequencing libraries 

- Sequencing via NGS

Haque et al. (2017). https://genomemedicine.biomedcentral.com/articles/10.1186/s13073-017-0467-4#citeas

2. Computational procedure (Depending on the package used)

This procedure is based on the Seurat package and the outputs from 10xGenomics.
This pipeline is adapted from the tutorial by Dr. Zhisong He and Prof. Barbara 
Treutlein at: https://github.com/quadbio/scRNAseq_analysis_vignette

- Import the package 

- Create the Seurat object by combining the 3 output files 

- Quality control

- Normalization 

- Feature selection for heterogeneity analysis 

- Scaling 

- Linear dimension reduction: Principal Component Analysis (PCA)

- Non-linear dimension reduction: t-distributed Stochastic Neighbor Embedding (t-SNE) and Uniform Manifold Approximation and Projection (UMAP) 

- Cell clustering and annotation


### 1) Project set up

```{r}
### Set working directory
setwd("/Users/manhduynguyen/STA330_ScRNA-Seq_PBMCs_10xGenomics")
```

```{r, eval = F}
### Install required packages
install.packages("leiden")
install.packages("Seurat")
install.packages("Matrix")
install.packages("patchwork")
```

```{r}
### Load packages
library("leiden")
library("Seurat")
library("Matrix")
library("ggplot2")
library("patchwork")
library("tidyverse")

if (!require("scCATCH")) {
  remotes::install_github("ZJUFanLab/scCATCH")
  library(scCATCH)
}

### List all loaded packages
(.packages())
```

### 2) Import datasets from 10xGenomics and create Seurat object

```{r}
### Import the raw scRNA-Seq data
# Expression matrix in sparse format
counts <- readMM("Raw Data/matrix.mtx.gz")
# Cell barcodes
barcodes <- read.table("Raw Data/barcodes.tsv.gz", 
                       stringsAsFactors = F) [,1]
# Gene features
features <- read.csv("Raw Data/features.tsv.gz", 
                     stringsAsFactors = F, 
                     sep = "\t", 
                     header = F)

### Check the structure of each imported file
str(counts)
str(barcodes)
str(features)

### Set gene features as row names and cell barcodes as column names
rownames(counts) <- make.unique(features[,2]) # Make sure that gene names are unique to eliminate duplicates
colnames(counts) <- barcodes

### Create Seurat object

# dgTMatrix (Triplet Matrix): A sparse matrix format that stores the matrix as a list
# of its non-zero values, along with their row and column indices. There are 3 components
# of values, row indices, and column indices

# dgCMatrix (Compressed Sparse Column Matrix): A sparse matrix that stores the matrix in 
# column-compressed form. The matrix is represented by 3 arrays of values, row indices
# of the values, and column pointers.

seurat <- CreateSeuratObject(counts, 
                             project = "5K_Mouse")

### Add mitochondrial content percentage as a new metadata column
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, 
                                               pattern = "^MT-|^mt-|^Mt-")

### Check the distribution
# Count per cell
summary(seurat[["nCount_RNA"]])
# Genes detected per cell
summary(seurat[["nFeature_RNA"]])
# Mitochondrial percentages
summary(seurat[["percent.mt"]])

### Check for NA values in each column of the metadata
### There should not be any NA (return value of 0)
sum(is.na(seurat@meta.data$nFeature_RNA))     
sum(is.na(seurat@meta.data$nCount_RNA))       
sum(is.na(seurat@meta.data$percent.mt))      

### Since there are 261743 NA values in the mitochondrial percentages, these NAs 
### should be removed
# Filter rows with NAs in either one of the three columns
filtered_meta_data <- seurat@meta.data %>%
  filter(!is.na(nFeature_RNA) & !is.na(nCount_RNA) & !is.na(percent.mt))

# Subset seurat object to only retain rows without NAs 
seurat_noNA <- subset(seurat, cells = rownames(filtered_meta_data))

# Make sure that there is no NAs in the column for mitochondrial percentages
sum(is.na(seurat_noNA@meta.data$percent.mt))

### Check the distribution again
### There should not be any NA (return value of 0)
# Count per cell
summary(seurat_noNA@meta.data$nCount_RNA)
# Genes detected per cell
summary(seurat_noNA@meta.data$nFeature_RNA)
# Mitochondrial percentages
summary(seurat_noNA@meta.data$percent.mt)

### nCount_RNA: The range is from 1 to 92216, with a median of 1 and a mean of 123.5, 
### suggesting that many cells are at the lower end of the scale, which is unusual 
### for single-cell RNA-seq data and might suggest low-quality cells.

### nFeature_RNA: The range is from 1 to 7825 features, with a median of 1 and a 
### mean of 41.31. suggesting that many cells have very few features detected, which 
### could be a sign of poor-quality or empty droplets.

### percent.mt: The minimum is 0, but the maximum is 100, suggesting that some cells 
### with a very high mitochondrial content, which is often associated with damaged 
### cells or dead cells.

### Create a violin plot with each point as an individual cell
# png("Plot_Violin__nCount_nFeature_percent.mt.png", res = 300, width = 8, height = 5, units = "in")
VlnPlot(seurat_noNA, features = c("nCount_RNA",
                                  "nFeature_RNA", 
                                  "percent.mt"), 
        ncol = 3)
# dev.off()

### Scatter plot to examine the relationship between RNA transcript count and 
### mitochondrial percentages
plot_scatter_nCountRNA_percentmt <- FeatureScatter(seurat_noNA, 
                                                   feature1 = "nCount_RNA", 
                                                   feature2 = "percent.mt",
                                                   raster = T)

plot_scatter_nCountRNA_percentmt + 
  geom_hline(yintercept = 5, linetype = "dashed", color = "black")

### Scatter plot to examine the relationship between RNA transcript count and 
### genes detected
plot_scatter_nCountRNA_nFeatureRNA <- FeatureScatter(seurat_noNA, 
                                                     feature1 = "nCount_RNA", 
                                                     feature2 = "nFeature_RNA",
                                                     raster = T)

plot_scatter_nCountRNA_nFeatureRNA + 
  geom_hline(yintercept = 500, linetype = "dashed", color = "black") + 
  geom_hline(yintercept = 5000, linetype = "dashed", color = "black")

### Combine the 2 scatter plots
# png("Plot_Relationship_nCount_nFeature_percent.mt.png", res = 300, width = 12, height = 5, units = "in")
plot_scatter_nCountRNA_percentmt + plot_scatter_nCountRNA_nFeatureRNA
# dev.off()

### As expected, the number of detected genes and the number of detected transcripts
### are correlated while the mitochondrial transcript percentage is not
```

### 3) Quality Controls

The most common quality control is to filter out:

1. Cells with too few detected genes: These cells are not sequenced deep enough for
reliable characterization.

2. Cells with too many detected genes: These cells may represent doublets or multiplets 
(i.e., two or more cells in the same droplets, therefore sharing the same barcode).

3. Cells with high mitochrondrial transcript percentages: These cells may represent
cells under stress (e.g., hypoxia) which produce a lot of mitochondria or high amount
of truncated mitochondrial transcripts.

As most of the scRNA-Seq experiments use oligo-T to capture mRNAs, mitochondrial 
transcripts should be relatively under-represented due to their lack of poly-A tails, 
but it is unavoidable that some mitochondrial transcripts are captured. There is 
also evidence that stable poly-A tails exist in some mitochondrial transcripts but 
serve as a marker for degradation. 

```{r}
### Filter out cells with detected genes between 500 and 5000, and mitochondrial 
### percentage lower than 5%
seurat_noNA_filtered_1 <- subset(seurat_noNA, 
                                nFeature_RNA > 500 & nFeature_RNA < 5000 & percent.mt < 5)
nrow(seurat_noNA_filtered_1@meta.data)
### Sometimes other thresholds and more quality controls need to be performed 

### Maybe a different number of detected genes (brains can range up to 12000 or as low as 200)
```

The tutorial from bioconductor with unfiltered 10xGenomics PBMCs only uses percent.mt
to filter out damage cells using median absolute deviation.

Oelen, R., de Vries, D. H., Brugge, et al. (2022). Single-cell RNA-sequencing of 
peripheral blood mononuclear cells reveals widespread, context-specific gene expression 
regulation upon pathogenic exposure. Nature Communications, 13(1). https://doi.org/10.1038/s41467-022-30893-5 

Oelen et al. (2022) suggests the cut-off threshold to be over 200 genes detected 
and mitochondiral percentage lower than 8% and 15% based on their batches.

```{r}
### Filter out cells with detected genes > 200, and mitochondrial percentage 
### lower than 8%
seurat_noNA_filtered_2 <- subset(seurat_noNA,
                                 nFeature_RNA > 200 & percent.mt < 5)
nrow(seurat_noNA_filtered_2@meta.data)

### Filter out cells with detected genes > 100
seurat_noNA_filtered_3 <- subset(seurat_noNA,
                                 nFeature_RNA > 100)
nrow(seurat_noNA_filtered_3@meta.data)

### Filter out cells with detected genes > 100, and mitochondrial percentage 
### lower than 5%
seurat_noNA_filtered_4 <- subset(seurat_noNA,
                                 nFeature_RNA > 100 & percent.mt < 5)
nrow(seurat_noNA_filtered_4@meta.data)
```

We decided to proceed with the initial thresholds of detected genes between 500 and 5000,
and mitochondrial percentage lower than 5% since the number of remaining cells aligns
with the results from the 10xGenomics pipeline (https://cf.10xgenomics.com/samples/cell-vdj/9.0.0/5k_Mouse_PBMCs_5p_gem-x_5k_Mouse_PBMCs_5p_gem-x/5k_Mouse_PBMCs_5p_gem-x_5k_Mouse_PBMCs_5p_gem-x_web_summary.html).

Xie, X., Liu, M., Zhang, Y., et al. (2020). Single-cell transcriptomic landscape of 
human blood cells. National Science Review, 8(3). https://doi.org/10.1093/nsr/nwaa180 

Stankiewicz, A. M., Goscik, J., Swiergiel, A. H., et al. (2014). Social stress increases 
expression of hemoglobin genes in mouse prefrontal cortex. BMC Neuroscience, 15(1). https://doi.org/10.1186/s12868-014-0130-6 

Xie et al. (2020) & Stankiewicz et al. (2014) suggest the marker genes for RBCs/erythrocytes: 
Hbb-b1, Hbb-b2, Hba-a1, Hba-a2, hbd, hbg, Beta-S, Alas2, Mgp, Gata1, Klf1, Klf3

```{r}
### Check for the potential presence of RBCs

# Define RBC-related genes
rbc_genes <- c("Hbb", "Hba1", "Hba2", "Hbd", "Gata1", "Klf1", "Klf3", 
               "Hbb-b1", "Hbb-b2", "Hba-a1", "Hba-a2", "hbd", "hbg", 
               "Beta-S", "Alas2", "Mgp")

# Extract gene names from the metadata of the filtered seurat 
genes_names_after_filtered_1 <- rownames(seurat_noNA_filtered_1)

# Check which RBC genes are present
genes_present <- rbc_genes %in% genes_names_after_filtered_1
print(genes_present)
```

### 4) Normalization

The amount of captured RNA is different from cell to cell. Therefore, it should be
avoided to directly compare the number of captured transcript for each gene between
cells. A normalization step, aiming to make gene expression levels between different 
cells comparable, is necessary.

The most commonly used normalization in scRNA-Seq analysis is to normalize the 
feature expression measurements for each cell to the total expression, and then
multiplies this by a scale factor or 10000 by default. The resulting expression 
levels are log-transformed so that the expression values better fit a normal 
distribution.

It is worth to mention that before doing the log-transformation, a pseudocount is 
added to every value so that genes with 0 transcripts detected in a cell still present
with value of 0 after log-transform.

Normalized Expression = log10(UMI counts/Total counts per cell*10^4 + 1)

```{r}
### Normalize the expression levels of all features
seurat_normalized_1 <- NormalizeData(seurat_noNA_filtered_1)

### Before normalization

# Extract the raw UMI count from the metadata as matrix before normalization
counts_matrix <- as.matrix(GetAssayData(seurat_noNA_filtered_1, layer = "counts"))

# Convert the raw count matrix into a numeric vector for visualization 
counts_vector <- as.numeric(counts_matrix)  

# Create the histogram of log-transformed raw UMI counts to assess data distribution
# The (log10(UMI counts + 1)) is the pseudocount for better interpretation
# png("Plot_hist_before_normalization.png", res = 300, width = 8, height = 5, units = "in")
hist(log10(counts_vector + 1), 
     breaks = 100, 
     main = "Before Normalization", 
     xlab = "log10(UMI counts + 1)")
# dev.off()

# png("Plot_hist_before_normalization_without_pseudo.png", res = 300, width = 8, height = 5, units = "in")
hist(log10(counts_vector), 
     breaks = 100, 
     main = "Before Normalization without Pseudocount", 
     xlab = "log10(UMI counts)")
# dev.off()

### After normalization

# Extract the raw UMI count from the metadata as matrix after normalization
# The "data" layer stores the log-normalized counts
normalized_matrix <- as.matrix(GetAssayData(seurat_normalized_1, layer = "data"))

# Convert the normalized count matrix into a numeric vector for visualization 
normalized_vector <- as.numeric(normalized_matrix)  # Convert to numeric vector

# Create the histogram of log-transformed normalized UMI counts to assess data distribution
# The (log10(UMI counts + 1)) is the pseudocount for better interpretation
# png("Plot_hist_after_normalization.png", res = 300, width = 8, height = 5, units = "in")
hist(log10(normalized_vector + 1), 
     breaks = 100, 
     main = "After Normalization", 
     xlab = "log10(Normalized counts + 1)")
# dev.off()

# png("Plot_hist_after_normalization_without_pseudo.png", res = 300, width = 8, height = 5, units = "in")
hist(log10(normalized_vector), 
     breaks = 100, 
     main = "After Normalization", 
     xlab = "log10(Normalized counts)")
# dev.off()
```


In scRNA-seq analysis, genes with zero counts in some or even most cells are not 
immediately removed during normalization. This is because:

1. Zero Counts May Be Due to Dropout Events
scRNA-seq data is highly sparse, meaning some genes are not detected even if they 
are actually expressed due to technical limitations (e.g., low capture efficiency).
Keeping these genes allows normalization methods to account for dropout events properly.

2. Some Genes Are Expressed in Only a Few Cells
A gene might have zero expression in most cells but be highly expressed in a rare 
cell type (e.g., a marker gene for a specific cell population). If removed prematurely, 
important cell-type-specific information could be lost.

3. Normalization Adjusts for Technical Variability
Normalization methods (like `NormalizeData()` in Seurat) scale gene expression 
relative to total counts per cell, ensuring fair comparison between genes.
Removing zero-expressed genes beforehand could bias this scaling process.

### 5) Feature Selection for Heterogeneity Analysis

The advantage of single-cell compared to bulk RNA-Seq is the potential to look into
cellular heterogeneity of samples, by identifying cell groups with distinct molecular 
signatures. However, not every gene has the same level of  information and the same
contribution when trying to identify different cell groups.

Therefore, it is necessary to perform a proper feature selection before further exploration.

```{r}
### Identify highly variable features/genes (genes with the most varied expression
### levels across cells)

# Top 5000 features
seurat_featurevariable_1 <- FindVariableFeatures(seurat_normalized_1,
                                                 nfeatures = 5000)
```

By default, Seurat calculates the standardized variance of each gene across cells,
and pick the top ones as the highly variable features. There is no good criteria to 
determine how many highly variable features to use. Sometimes it might require a few
iterations to pick the number that gives the clearest for interpretation. A value 
between 2000 and 5000 should be good enough.

```{r}
### Extract the top 20 features that are highly variable
top_20_features <- head(VariableFeatures(seurat_featurevariable_1), 20)
print(top_20_features)

### Plot the variable features 

# Create a scatter plot of feature variance to visualize the distribution of variable 
# genes across dataset
plot_topfeatures_1 <- VariableFeaturePlot(seurat_featurevariable_1)

# Annotate the top 20 variables on the scatter plot
plot_topfeatures_anno_1 <- LabelPoints(plot = plot_topfeatures_1, 
                                       points = top_20_features,
                                       repel = T,
                                       xnudge = 0,
                                       max.overlaps = 25)

# png("Plot_featureselection.png", res = 300, width = 8, height = 5, units = "in")
plot_topfeatures_anno_1
# dev.off()
```


### 5) Data Scaling

Since different genes have different base expression levels and distributions,
the contribution of each gene to the analysis is different if no data transformation
is performed. These highly expressed genes/features would dominate the dimension 
reduction techniques in the later steps of the analysis. Therefore, we have to scale 
the data with the selected features.

The scaling is performed using z-score normalization on the log-normalized data,
so that the expression of each gene centers around 0, with an SD of 1. 

Scaled Expression = (Normalized Expression - Mean of Normalized Expression)/SD of Normalize Expression

During this step, it is possible to remove unwanted sources of variation from 
the dataset such as nFeature_RNA/nCount_RNA, percent.mt, or the cell cycle related 
variables.

A common suggestion is to not perform any removal in the first iteration of data
exploration but to check the results first and if any unwanted source of variation
dominates cellular heterogeneity, try to regress out the respective variable and see
whether the result improves.

```{r}
### Normalize the data using selected features
seurat_scaled_1 <- ScaleData(seurat_featurevariable_1)
```


```{r}
### Create plot for expression distribution to compare the log-normalized data and the scaled data

# Get top 3 variable genes 
top_variable_genes <- head(VariableFeatures(seurat_scaled_1), 3)

# Extract log-normalized and scaled expression values for filtered genes
normalized_data <- GetAssayData(seurat_scaled_1, slot = "data")[top_variable_genes, ]
scaled_data <- GetAssayData(seurat_scaled_1, slot = "scale.data")[top_variable_genes, ]

# Extract the log-normalized expression values for these genes
normalized_data <- GetAssayData(seurat_scaled_1, slot = "data")[top_variable_genes, ]

# Extract the scaled expression values for these genes
scaled_data <- GetAssayData(seurat_scaled_1, slot = "scale.data")[top_variable_genes, ]

# Create an empty list to store the plots
plot_list <- list()

# Loop through each top variable gene/feature for density plots
for (gene in top_variable_genes) {
  # Create a data frame of log-normalized and scaled expression values fro each gene 
  df <- data.frame(
    Expression = c(normalized_data[gene, ], scaled_data[gene, ]),
    Type = rep(c("Log-Normalized", "Scaled"), each = ncol(normalized_data)),
    Gene = gene
  )
  
  # Create the density plots based on the extracted expression values
  plot_list[[gene]] <- ggplot(df, aes(x = Expression, fill = Type)) +
    geom_density(alpha = 0.7) +
    labs(y = "Density", 
         title = gene) +
    theme_minimal()
}

# Combine and display the plots
# png("Plot_scale_top3variablegene.png", res = 300, width = 8, height = 5, units = "in")
wrap_plots(plot_list, ncol = 1) +
  plot_annotation(title = "Expression Distribution Before vs After Scaling")
# dev.off()

# Compared the log-normalized and scaled data for the expression of the top 3 variable genes
summary(as.vector(GetAssayData(seurat_scaled_1, slot = "data")["S100a9", ]))
summary(as.vector(GetAssayData(seurat_scaled_1, slot = "scale.data")["S100a9", ]))

summary(as.vector(GetAssayData(seurat_scaled_1, slot = "data")["S100a8", ]))
summary(as.vector(GetAssayData(seurat_scaled_1, slot = "scale.data")["S100a8", ]))

summary(as.vector(GetAssayData(seurat_scaled_1, slot = "data")["Igkv6-15", ]))
summary(as.vector(GetAssayData(seurat_scaled_1, slot = "scale.data")["Igkv6-15", ]))
```

Log-normalized data:

- Data have many 0s, especially for sparse genes 

- The mean is > 0, due to a small number of cells with high expression pulling it up

- The median and quartiles are 0 as most cells do not express genes

This distribution is not symmetric and not centered around 0, as expected of log-normalized
data

Scaled data:

- The mean is centered at 0 

- Since there are so many 0s that all of the quartiles have the same negative z-scores

The sparse nature cause the scaled data to be extremely right skewed

### 6) Principal Component Analysis (PCA)

Cell heterogeneity can be described after identifying highly variable genes and 
scaling the data. However, applying a linear dimension reduction with PCA is recommended
as:

1. The data become more compact so the computation becomes faster

2. As scRNA-Seq data is intrinsically sparse, summarizing the measurements of related
features greatly enhances the signal robustness

The number of PCs that can be calculated for a dataset is equal to the number of highly 
variable features/genes or the number of cells, whichever value is smaller. However,
most of these PCs are not informative and only represent random noise. Only the top
PCs are informative and represent differences among cell populations. Therefore, instead
of calculating all possible PCs, Seurat uses truncated PCA to only calculate the first 
50 PCs by default. 

In order to pick the optimal number of PCs, we can use the elbow plot. Higher ranked PCs
(have higher standard deviations) explain more variation in the data than lower-ranked 
PCs. However, the decrease in standard deviation is not linear. The curve of the elbow
plot drops dramatically for the first few PCs and slowly flatten out. This would assume 
that the first phase of the curve represents the real signal related to biological 
differences between cell populations, while the second phase represents mostly technical
variation or the stochastic nature of individual cells.

In addition, we can also check which genes are mostly contributing to each of the 
top PCs. This can be informative if we know the genes and the biology of the analyzed
samples. It provides the opportunity to understand the biological implication of each
of the top PCs

```{r}
### Perform the default PCA with 50 first PCs
seurat_pca_1 <- RunPCA(seurat_scaled_1, npcs = 50)

### Visualize the PCA using scatter plot
DimPlot(seurat_pca_1, reduction = "pca")

### Create the elbow plot to identify the optimal number of PCs
# png("Plot_PCElbow.png", res = 300, width = 8, height = 5, units = "in")
ElbowPlot(seurat_pca_1, ndims = ncol(Embeddings(seurat_pca_1, "pca")))
# dev.off()

# The optimal PCs seems to be 10

### Check the top 10 PCs with heatmap
# png("Plot_PCHeatmap.png", res = 300, width = 8, height = 5, units = "in")
PCHeatmap(seurat_pca_1, 
          dims = 1:10, 
          cells = 500, 
          balanced = T,
          ncol = 5)
# dev.off()
```

It is not recommended to choose only PCs represented by "interesting" genes. For 
most of the data, a PC number from 10 to 50 would be reasonable.

### 7) t-distributed Stochastic Neighbor embedding (t-SNE) and Uniform Manifold Approximation and Projection (UMAP)

Both methods perform non-linear dimensional reduction, aiming to map each cell 
into a lower-dimensional space (2D/3D) while preserving the original distances or 
neighborhood relationships as accurately as possible.

Both methods have their pros and cons and neither always work better than the other.
t-SNE provides great visualization when cells form distinct cell groups while UMAP
preserves trajectory-like structure better when data contain "continuum" (e.g. the 
continuous cell state change during development and differentiation) 

```{r}
### Perform t-SNE and UMAP

# The top PCs from the PCA are used as input to create the t-SNE and UMAP embedding 
# of the data 
seurat_tSNE_1 <- RunTSNE(seurat_pca_1, dims = 1:10)
seurat_UMAP_1 <- RunUMAP(seurat_pca_1, dims = 1:10)

# Create a scatter plot of the t-SNE and UMAP embeddings
plot_tSNE <- TSNEPlot(seurat_tSNE_1)
plot_UMAP <- UMAPPlot(seurat_UMAP_1)

# Display both plots side by side for comparison 
# png("Plot_tSNE_nocluster.png", res = 300, width = 8, height = 5, units = "in")
plot_tSNE
# dev.off()

# png("Plot_UMAP_nocluster.png", res = 300, width = 8, height = 5, units = "in")
plot_UMAP
# dev.off()
```

Once t-SNE and UMAP embedding is created, we can start checking whether certain 
cell types or cell states exist in the data by doing feature plots of some known
canonical markers of the cell types of interest

```{r}
### Visualize the expression of marker genes on the t-SNE and UMAP embedding

### Known marker genes
# T-cells: Cd4, Cd8a, Cd8b, Cd3e
# B-cells: Cd19, Cd79a, Cd79b
# Natural killer cells: Ncam1 (Cd56), Klrd1 (Cd94)
# Monocytes: Cd14, Fcgr3a (Cd16)

# t-SNE
plot_tSNE_genes <- FeaturePlot(seurat_tSNE_1, c("Cd4", "Cd8a", "Cd3e", "Cd19", 
                                                "Cd79a", "Cd79b", "Ncam1", "Krld1", 
                                                "Cd14", "Fcgr3a"),
                               ncol = 4, 
                               reduction = "tsne")

# UMAP
plot_UMAP_genes <- FeaturePlot(seurat_UMAP_1, c("Cd4", "Cd8a", "Cd3e", "Cd19", 
                                                "Cd79a", "Cd79b", "Ncam1", "Krld1", 
                                                "Cd14", "Fcgr3a"),
                               ncol = 4, 
                               reduction = "umap")

# Display both plots side by side for comparison
# png("Plot_tSNE_genes.png", res = 300, width = 14, height = 5, units = "in")
plot_tSNE_genes
# dev.off()

# png("Plot_UMAP_genes.png", res = 300, width = 14, height = 5, units = "in")
plot_UMAP_genes
# dev.off()
```

### 8) Cell Clustering

Creating feature plot of known markers is usually a good way to start with when 
exploring scRNA-seq data. However, to more comprehensively understand the underlying 
heterogeneity in the data, it is necessary to identify cell groups with an unbiased 
manner, resulting in the use of clustering. 

(t-SNE/UMAP only provide visualization of high-dimensional data in 2D, clustering 
provides a quantitative identification of distinct cell population)

In theory, any clustering methods can be used, including those widely used in bulk 
RNA-seq data analysis such as hierarchical clustering and k-means. However, in practice, 
this is very difficult, as the sample size in scRNA-seq data is much larger (one 
10x experiment usually gives several thousands of cells). It would be extremely 
slow to use these methods. In addition, due to the intrinsic sparseness of scRNA-seq 
data, even if data is denoised by dimension reduction like PCA, differences between 
different cells are not as well quantitative as those of bulk RNA-seq data. Therefore, 
the more commonly used clustering methods in scRNA-seq data analysis is graph-based 
community identification algorithm. 

First, a k-nearest neighbor (kNN) network of cells is constructed. Each cell is 
initially connected to its closest neighbors based on their PC values. Only pairs 
of cells that are mutual neighbors are considered connected. The proportion of 
shared neighbors between each cell pair is then calculated to determine the strength 
of their connection, with weak connections being pruned. The result is a Shared 
Nearest Neighbor (SNN) network.

Once the network is built, the Louvain community detection algorithm is applied 
to identify clusters within the network. This algorithm groups cells into communities 
where intra-cluster connections are dense, while inter-cluster connections are sparse, 
effectively identifying biologically relevant cell populations.


```{r}
### Perform clustering and visualize on t-SNE/UMAP embedding

# Construct kNN graph 
seurat_tSNE_neighbor_1 <- FindNeighbors(seurat_tSNE_1, dims = 1:10)

# Cluster cells using Louvain algorithm 
seurat_tSNE_cluster_1 <- FindClusters(seurat_tSNE_neighbor_1, resolution = 1)

# Visual the clusters on t-SNE embedding 
plot_tSNE_cluster_1 <- DimPlot(seurat_tSNE_cluster_1, reduction = "tsne", label = T)

# Construct kNN graph 
seurat_UMAP_neighbor_1 <- FindNeighbors(seurat_UMAP_1, dims = 1:10)

# Cluster cells using Louvain algorithm 
seurat_UMAP_cluster_1 <- FindClusters(seurat_UMAP_neighbor_1, resolution = 1)

# Visual the clusters on UMAP embedding 
plot_UMAP_cluster_1 <- DimPlot(seurat_UMAP_cluster_1, reduction = "umap", label = T)

# Display both plots side by side for comparison
plot_tSNE_cluster_1 + plot_UMAP_cluster_1

png("Plot_tSNE_cluster.png", res = 300, width = 14, height = 5, units = "in")
plot_tSNE_cluster_1
dev.off()

png("Plot_UMAP_cluster.png", res = 300, width = 14, height = 5, units = "in")
plot_UMAP_cluster_1
dev.off()
```

The `resolution` parameter in `FindClusters()` determines the granularity of clustering. 
A lower resolution (e.g., 0.1) yields larger, coarse-grained clusters, representing 
broad cell types, while a higher resolution (e.g., 1) produces smaller, more refined 
clusters, capturing finer cell subtypes. Commonly used values range between 0.1 and 1, 
depending on the dataset and desired clustering detail.


### 9) Cluster Annotation

After clustering, cluster annotation assigns biological identities to each cluster.
This process involves several integrated steps

- Identify marker genes that are significantly enriched in each cluster
(`FindAllMarkers()` function performs differential expression analysis with Wilcoxon's
rank sum test between cells in the cluster and cells in other clusters. Due to the intrinsic
larger sample size in scRNA-Seq data, with one cell as one sample, it is strongly recommended
to not look at only p-value, but also the minimum detection rate of the gene in the cluster
`min.pct` and the log fold change between cells in and outside the cluster `logfc.threshold`)

In the context of scRNA-seq, differential expression analysis (e.g., using the Wilcoxon 
rank sum test) compares the distribution of gene expression levels for each gene in 
the cluster of interest against the distribution in cells from other clusters, as 
determined by the prior clustering step. This helps identify marker genes that are 
significantly more (or less) expressed in a given cluster compared to others.

- Compare the identified cluster markers to curated reference datasets (e.g. CellMatch)
to determine the overlap between the cluster marker genes and known cell type marker
genes

- Automatically predict the preliminary cell type annotations based on the overlapping 
marker genes using the `scCATCH` package

- Create a curated list of key marker genes from references and identify the the gene 
expression pattern of these marker genes across all clusters

- Manually adjusted the preliminary cell type annotations based on the gene expression 
patterns using the curated marker genes

- Visualize the cell type annotations on the t-SNE and UMAP embeddings

```{r}
# Define a function to analyze markers that overlap with reference genes
analyze_cluster_markers <- function(object) {
  # Extract all unique cluster ID from the "markergene" slot of the object 
  clusters <- unique(object@markergene$cluster)
  
  # Loop through each cluster to compare the marker genes and reference genes 
  for (cluster_id in clusters) {
    # Get unique marker genes for the current cluster 
    cluster_markers <- unique(object@markergene$gene[object@markergene$cluster == cluster_id])
    
    # Find the overlapping genes between the marker genes and reference genes in the current cluster
    overlapping_genes <- intersect(cluster_markers, unique(object@marker$gene))
    
    # Print the cluster ID
    cat(paste0("\nCluster: ", cluster_id, "\n"))
    
    # Print the number of marker genes in the cluster
    cat(paste0("  Number of marker genes: ", length(cluster_markers), "\n"))
    
    # Print the number of overlapping genes
    cat(paste0("  Number of overlapping genes with reference: ", length(overlapping_genes), "\n"))
    
    # If there are overlapping genes, print the first 10; otherwise, state that no overlaps were found 
    if (length(overlapping_genes) > 0) {
      cat("  Overlapping genes (first 10):\n")
      print(head(overlapping_genes, 10))
    } else {
      cat("  No overlapping genes found with the reference.\n")
    }
  }
}

### t-SNE

# Identify top markers per cluster
sc_tSNE_markers <- FindAllMarkers(
  seurat_tSNE_cluster_1,
  assay = "RNA",
  only.pos = TRUE,
  min.pct = 0.25,        # Detection rate the gene in the cluster
  logfc.threshold = 0.5, # The fold change between cells in and out the cluster
  test.use = "wilcox"
)

# Check the top 2 marker genes per cluster based on avg_log2FC
sc_tSNE_marker_top <- sc_tSNE_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 2) %>%
  as.data.frame()

# Create scCATCH object using raw counts and cluster IDs
sc_tSNE_obj <- createscCATCH(
  data = seurat_tSNE_cluster_1@assays$RNA$counts, # Lost dimnames when use seurat_tSNE_cluster_1@assays$RNA@layers$counts
  cluster = as.character(Idents(seurat_tSNE_cluster_1))
)

# Add marker genes and reference database based on the species and tissue types
data("cellmatch")  
sc_tSNE_obj@markergene <- sc_tSNE_markers
sc_tSNE_obj@marker <- cellmatch[cellmatch$species == "Mouse" &
                                cellmatch$tissue %in% c("Peripheral blood", "Blood"), ]

# Analyze markers overlapping with the reference genes
analyze_cluster_markers(sc_tSNE_obj)

# Predict cell types 
sc_tSNE_obj <- findcelltype(sc_tSNE_obj)

# Check the predicted cell types with scores
sc_tSNE_obj@celltype %>%
  select(cluster, cell_type, celltype_score)


### UMAP

# Identify top markers per cluster
sc_UMAP_markers <- FindAllMarkers(
  seurat_UMAP_cluster_1,
  assay = "RNA",
  only.pos = TRUE,
  min.pct = 0.25,         # Detection rate the gene in the cluster
  logfc.threshold = 0.5,  # The fold change between cells in and out the cluster
  test.use = "wilcox"
)

# Check the top 2 marker genes per cluster based on avg_log2FC
sc_UMAP_marker_top <- sc_UMAP_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 2) %>%
  as.data.frame()

# Create scCATCH object using raw counts and cluster IDs
sc_UMAP_obj <- createscCATCH(
  data = seurat_UMAP_cluster_1@assays$RNA$counts, # Lost dimnames when use seurat_UMAP_cluster_1@assays$RNA@layers$counts
  cluster = as.character(Idents(seurat_UMAP_cluster_1))
)

# Add marker genes and reference database based on the species and tissue types
data("cellmatch")  
sc_UMAP_obj@markergene <- sc_UMAP_markers
sc_UMAP_obj@marker <- cellmatch[cellmatch$species == "Mouse" &
                                cellmatch$tissue %in% c("Peripheral blood", "Blood"), ]

# Analyze markers overlapping with the reference genes
analyze_cluster_markers(sc_UMAP_obj)

# Predict cell types 
sc_UMAP_obj <- findcelltype(sc_UMAP_obj)

# Check the predicted cell types with scores
sc_UMAP_obj@celltype %>%
  select(cluster, cell_type, celltype_score)
```

The curated marker gene list is created based on these references: 

Hackert, N. S., Radtke, F. A., Exner, T., et al. (2023). Human and mouse neutrophils 
share core transcriptional programs in both homeostatic and inflamed contexts. 
Nature Communications, 14(1). https://doi.org/10.1038/s41467-023-43573-9 

Miyake, K., Ito, J., Nakabayashi, J., Shichino, S., Ishiwata, K., & Karasuyama, H. (2023). 
Single cell transcriptomics clarifies the Basophil differentiation trajectory and identifies pre-basophils upstream of mature basophils. Nature Communications, 14(1). https://doi.org/10.1038/s41467-023-38356-1 

Wang, Z., Xie, L., Ding, G., et al. (2021). Single-cell RNA sequencing of peripheral 
blood mononuclear cells from Acute Kawasaki Disease patients. Nature Communications, 12(1). https://doi.org/10.1038/s41467-021-25771-5 

```{r}
# Define the curated cell type markers
curated_markers <- list()
curated_markers[["T-cells"]] <- c("Cd3d", "Cd3e", "Cd3g")
curated_markers[["CD4+ T-cells"]] <- c("Cd4")
curated_markers[["CD8+ T-cells"]] <- c("Cd8a", "Cd8b")
curated_markers[["NK-cells"]] <- c("Ncam1", "Klrb1", "Nkg7")
curated_markers[["B-cells"]] <- c("Cd19", "Ms4a1", "Cd38")
curated_markers[["Monocytes"]] <- c("Cd14", "Cd68", "Cd16")
curated_markers[["Neutrophil"]] <- c("Csf3r", "Cxcr2", "Ncf4", "Junb")
curated_markers[["Basophil"]] <- c("Fcer1a", "Clec12a", "Cd9", "Itga2")
curated_markers[["mDCs"]] <- c("Cd1c") # Myeloid dendritic cells
curated_markers[["pDCs"]] <- c("Lilra4") # Plasmacytoid dendritic cells 
curated_markers[["HSPCs"]] <- c("Cd34") # Hematopoietic stem and progenitor cells 
curated_markers[["Erythrocytes"]] <- c("Hba-a1", "Hba-a2", "Hbb-b1", "Hbb-b2")
curated_markers[["Megakaryocytes"]] <- c("Ppbp")

# Create dot plot for the expression level of each curated gene in each cluster 
sc_tSNE_markers_dotplot <- DotPlot(seurat_tSNE_cluster_1, 
                              features = curated_markers, 
                              assay = 'RNA') +
  theme(text = element_text(size=10), 
        axis.text.x = element_text(angle = 45, vjust = 0.5))


sc_UMAP_markers_dotplot <- DotPlot(seurat_UMAP_cluster_1, 
                              features = curated_markers, 
                              assay = 'RNA') +
  theme(text = element_text(size=10), 
        axis.text.x = element_text(angle = 45, vjust = 0.5))


# Display the plots
# These two should be similar
# png("Plot_tSNE_markergenes_expression.png", res = 300, width = 24, height = 5, units = "in")
sc_tSNE_markers_dotplot
# dev.off()

# png("Plot_UMAP_markergenes_expression.png", res = 300, width = 24, height = 5, units = "in")
sc_UMAP_markers_dotplot
# dev.off()

# Genes not found: Cd8b, Cd16, Cd1c, Lilra4, Hbb-b1, Hbb-b2

# Based on these gene patterns
# Cluster 3: CD8+ => CD4+
# Cluster 7: Stem cell => Unknown 1
# Cluster 8: HSPCs => Unknown 2
# Cluster 9: Regulatory => CD4+
# Cluster 11: CD8+ => NK
# Cluster 12: Regulatory => Unknown 3
# Cluster 14: Basophil => Basophil/Neutrophil
# Cluster 15: Platelet => Basophil/Megakaryocyte

### t-SNE

# Create a new dataframe for cell type annotation
sc_tSNE_anno <- sc_tSNE_obj@celltype %>%
  select(cluster, cell_type) %>%
  mutate(cluster = factor(cluster, levels = c(0:15))) %>%
  arrange(cluster)

# Check the new dataframe
sc_tSNE_anno

# Update the annotations for cell cluster based on the dot plot and scCATCH results
sc_tSNE_anno$cell_type[c(4,10)] <- "CD4+ T Cell"
sc_tSNE_anno$cell_type[c(8)] <- "Unknown 1"
sc_tSNE_anno$cell_type[c(9)] <- "Unknown 2"
sc_tSNE_anno$cell_type[c(12)] <- "Natural Killer Cell"
sc_tSNE_anno$cell_type[c(13)] <- "Unknown 3"
sc_tSNE_anno$cell_type[c(15)] <- "Basophil/Neutrophil"
sc_tSNE_anno$cell_type[c(16)] <- "Basophil/Megakaryocyte"

# Extract the meta data from the seurat object
old_tSNE_metadata <- seurat_tSNE_cluster_1@meta.data

# Combine the cell type annotation into the meta data 
new_tSNE_metadata <- old_tSNE_metadata %>%
  left_join(sc_tSNE_anno, by = c("seurat_clusters" = "cluster"))

# Retain the original row names 
rownames(new_tSNE_metadata) <- rownames(seurat_tSNE_cluster_1@meta.data)

# Import the new meta data to the seurat object 
seurat_tSNE_cluster_1@meta.data <- new_tSNE_metadata

# Create the new t-SNE plot with cluster annotations
sc_tSNE_dimplot <- DimPlot(seurat_tSNE_cluster_1, 
                           group.by = "cell_type", 
                           reduction = "tsne",
                           label = T) +
  theme(text = element_text(size = 10))

### UMAP

# Create a new dataframe for cell type annotation
sc_UMAP_anno <- sc_UMAP_obj@celltype %>%
  select(cluster, cell_type) %>%
  mutate(cluster = factor(cluster, levels = c(0:15))) %>%
  arrange(cluster)

# Check the new dataframe
sc_UMAP_anno

# Update the annotations for cell cluster based on the dot plot and scCATCH results
sc_UMAP_anno$cell_type[c(4,10)] <- "CD4+ T Cell"
sc_UMAP_anno$cell_type[c(8)] <- "Unknown 1"
sc_UMAP_anno$cell_type[c(9)] <- "Unknown 2"
sc_UMAP_anno$cell_type[c(12)] <- "Natural Killer Cell"
sc_UMAP_anno$cell_type[c(13)] <- "Unknown 3"
sc_UMAP_anno$cell_type[c(15)] <- "Basophil/Neutrophil"
sc_UMAP_anno$cell_type[c(16)] <- "Basophil/Megakaryocyte"

# Extract the meta data from the seurat object
old_UMAP_metadata <- seurat_UMAP_cluster_1@meta.data

# Combine the cell type annotation into the meta data 
new_UMAP_metadata <- old_UMAP_metadata %>%
  left_join(sc_UMAP_anno, by = c("seurat_clusters" = "cluster"))

# Retain the original row names 
rownames(new_UMAP_metadata) <- rownames(seurat_UMAP_cluster_1@meta.data)

# Import the new meta data to the seurat object 
seurat_UMAP_cluster_1@meta.data <- new_UMAP_metadata

# Create the new t-SNE plot with cluster annotations
sc_UMAP_dimplot <- DimPlot(seurat_UMAP_cluster_1, 
                           group.by = "cell_type", 
                           reduction = "umap",
                           label = T) +
  theme(text = element_text(size = 10))

# Display the plots
png("Plot_tSNE_final.png", res = 300, width = 14, height = 5, units = "in")
sc_tSNE_dimplot
dev.off()

png("Plot_UMAP_final.png", res = 300, width = 14, height = 5, units = "in")
sc_UMAP_dimplot
dev.off()

```

### 10) Integration into the meta-analysis project of chronic social defeat stress

```{r}
### Visualize the downregulated genes in both UMAP and tSNE 

# Sin3a, Tmem223, Tecr, Wipi2, Cd37, Gne, Mtss1, Wdr24, Npepl1, Pssmd7, Vars1, Rlig1
# Fxn, Akip1, Gcn1, Oard1, Ints11, Phf13, H2-Ab1, Fcho1, Gtf3c2, Isoc2b, Cep164


eli_plot_tSNE_down_genes <- FeaturePlot(seurat_tSNE_1, c("Sin3a", "Tmem223", "Tecr", "Wipi2", 
                                                         "Cd37", "Gne", "Mtss1", "Wdr24", 
                                                         "Npepl1", "Pssmd7", "Vars1", "Rlig1",
                                                         "Fxn", "Akip1", "Gcn1", "Oard1",
                                                         "Ints11", "Phf13", "H2-Ab1",
                                                         "Fcho1", "Gtf3c2", "Isoc2b", "Cep164"),
                                                         ncol = 4, 
                                                         reduction = "tsne")

eli_plot_tSNE_down_genes

eli_plot_UMAP_down_genes <- FeaturePlot(seurat_UMAP_1, c("Sin3a", "Tmem223", "Tecr", "Wipi2", 
                                                         "Cd37", "Gne", "Mtss1", "Wdr24", 
                                                         "Npepl1", "Pssmd7", "Vars1", "Rlig1",
                                                         "Fxn", "Akip1", "Gcn1", "Oard1",
                                                         "Ints11", "Phf13", "H2-Ab1",
                                                         "Fcho1", "Gtf3c2", "Isoc2b", "Cep164"),
                                                          ncol = 4, 
                                                          reduction = "umap")

eli_plot_UMAP_down_genes

### Visualize the upregulated genes in both UMAP and tSNE 

# St3gal5, Anks1b, Otos, Stradb, Guca2a, Cyp2e1, 4933411K16Rik?, Stip1, Pxmp2, Gpatch8
# Isg20, Rhbdl1, C1ql2, Nol4l, Scgb1a1, Arf5

eli_plot_tSNE_up_genes <- FeaturePlot(seurat_tSNE_1, c("St3gal5", "Anks1b", "Otos", "Stradb", 
                                                       "Guca2a", "Cyp2e1", "4933411K16Rik", 
                                                       "Stip1", "Pxmp2", "Gpatch8", "Isg20", "Rhbdl1",
                                                        "C1ql2", "Nol4l", "Scgb1a1", "Arf5"),
                                                        ncol = 4, 
                                                        reduction = "tsne")

eli_plot_tSNE_up_genes

eli_plot_UMAP_up_genes <- FeaturePlot(seurat_UMAP_1, c("St3gal5", "Anks1b", "Otos", "Stradb", 
                                                       "Guca2a", "Cyp2e1", "4933411K16Rik", 
                                                       "Stip1", "Pxmp2", "Gpatch8", "Isg20", "Rhbdl1",
                                                       "C1ql2", "Nol4l", "Scgb1a1", "Arf5"),
                                                       ncol = 4, 
                                                       reduction = "umap")

eli_plot_UMAP_up_genes


### Create a list of genes of interest

downregulate_genes <- c("Sin3a", "Tmem223", "Tecr", "Wipi2", "Cd37", "Gne", "Mtss1", "Wdr24", 
                        "Npepl1", "Pssmd7", "Vars1", "Rlig1", "Fxn", "Akip1", "Gcn1", "Oard1",
                        "Ints11", "Phf13", "H2-Ab1", "Fcho1", "Gtf3c2", "Isoc2b", "Cep164")

upregulate_genes <- c("St3gal5", "Anks1b", "Otos", "Stradb", "Guca2a", "Cyp2e1",
                      "4933411K16Rik", "Stip1", "Pxmp2", "Gpatch8", "Isg20", "Rhbdl1",
                      "C1ql2", "Nol4l", "Scgb1a1", "Arf5")

### Check if the interested genes are available in the original meta data 
setdiff(downregulate_genes,
        rownames(seurat_UMAP_cluster_1))
# Genes not found Pssmd7, Vars1, Rlig1

setdiff(upregulate_genes,
        rownames(seurat_UMAP_cluster_1))
# All genes found

# However, most of the these genes are not in the feature variables for scaling 
# (Only the top 5000 most variable features are included)

### Create the heatmaps with available genes
DoHeatmap(subset(seurat_UMAP_cluster_1, downsample = 100), 
          features = downregulate_genes, 
          group.by = "cell_type",
          size = 3)

DoHeatmap(subset(seurat_UMAP_cluster_1, downsample = 100), 
          features = upregulate_genes, 
          group.by = "cell_type",
          size = 3)

# Combine the 2 lists of genes together 
eli_gene_of_interest <- union(downregulate_genes, upregulate_genes)

# Check if the genes of interest are available
eli_gene_of_interest <- intersect(eli_gene_of_interest, rownames(seurat_normalized_1))

# Scale these genes of interests instead of the top 5000 variable features
eli_seurat_UMAP_cluster <- ScaleData(seurat_normalized_1, features = eli_gene_of_interest)

# Check if the cell barcodes match between the two objects
identical(colnames(eli_seurat_UMAP_cluster), colnames(seurat_UMAP_cluster_1))
# This should return TRUE

# Copy the cell_type annotation over from the UMAP object for heatmap
eli_seurat_UMAP_cluster@meta.data$cell_type <- seurat_UMAP_cluster_1@meta.data$cell_type

# Create the heatmap with these genes of interest
DoHeatmap(eli_seurat_UMAP_cluster,
          features = eli_gene_of_interest,
          group.by = "cell_type",
          size = 3)
```
